\section{Conclusion}
\label{sec:conclusion}
We introduced a three-route framework (\cref{sec:framework}) for disentangling representation and execution in algorithmic reasoning. The intermediate route, code generation followed by LLM simulation, lets us compare natural-language reasoning and code execution without changing both the trace representation and the executor at once. On the 40 algorithmic benchmark tasks, Route~1 and Route~2 are statistically close: Route~2 exceeds Route~1 by +0.15pp with a 95\% cluster-bootstrap CI of $[-0.30,+0.61]$pp. Route~3 reaches 48.84\% accuracy and exceeds Route~2 by +31.47pp with CI $[+29.20,+33.71]$pp (\cref{sec:performance}). The representation analysis in \cref{sec:rep_analysis} explains the small Route~1/Route~2 gap through a shared-core nuisance model and a reconstruction intervention showing that code-derived natural-language traces retain comparable downstream utility. The execution analysis in \cref{sec:route2v3} explains the large Route~3 gain through low recovery mass (1.61\%) and high execution-win mass (33.08\%, \cref{fig:recovery_final}). Taken together, these results suggest that code helps primarily by generating an executable trace that can be run reliably. In this setting, the central advantage of tool use is not the code representation, but the ability to hand the generated trace to a deterministic executor.


% Our causal intervention experiments (\cref{sec:rep_analysis}) demonstrate that NL reasoning traces are distributionally and functionally equivalent to code-translated traces, supporting the hypothesis that NL reasoning implicitly simulates underlying algorithmic computations for specific models, tasks and prompts. Theoretically, we prove that code-based reasoning achieves lower Bayes risk via Blackwell dominance (\cref{prop:dominance}), providing an information-theoretic explanation for the empirical advantages of solver-based pipelines.

\paragraph{Limitations.}
We highlight a few limitations. First, the main experiments do not fully cover frontier reasoning models. We ran controlled-subset frontier-model ablations (\Cref{tab:appendix_frontier_nopatch}) and found that the results generally hold. Second, task coverage is limited to algorithmic problems with externally verifiable answers and reliable Python execution, so open-ended, ambiguous, unsafe, or weakly specified tool use may behave differently. Third, the theory mainly builds intuition through standard decision-theoretic ideas and derives its novelty from the application, rather than providing a new general theorem about LLMs or code. Fourth, the linear models and shared-core/sufficient-statistic assumptions are motivated by practice but may fail when code omits decision-relevant information, natural language carries useful signal, or executors exploit structure outside the modeled core.


% \paragraph{Limitations}

% \paragraph{Future Work}

\section{Impact Statement}
This paper presents work whose goal is to advance the field of machine learning. There are many potential societal consequences of our work, none of which we feel must be specifically highlighted here.

\section*{Acknowledgements}
This work was partially funded by ONR Contract N00014-23-1-2417. We are grateful for resources and computational support provided by the Cognitive Computation Group at the University of Pennsylvania. We also thank the reviewers for their thoughtful feedback and comments.
