% Polished LaTeX appendix source included by example_paper.tex.
% Required packages in the main preamble: booktabs, array, caption, url, hyperref, cleveref.

\clearpage
\onecolumn

\captionsetup[table]{font=small,labelfont=bf,textfont=normalfont,labelsep=colon,skip=5pt,hypcap=false}

% Appendix-local notation and formatting helpers.
\newcommand{\RouteOne}{Route~1}
\newcommand{\RouteTwo}{Route~2}
\newcommand{\RouteThree}{Route~3}
\newcommand{\appendixresult}[1]{\par\noindent\textbf{Result.}~#1\par}
\newenvironment{appendixtableblock}{\par\smallskip\noindent\begin{minipage}{\linewidth}\centering}{\end{minipage}\par\smallskip}

\section{Supplementary Route Results}
\label{app:route_robustness}

\noindent\textbf{Route definitions.}~Across all route tables, \RouteOne{} denotes natural-language reasoning,
\RouteTwo{} denotes natural-language simulation over a code trace, and
\RouteThree{} denotes deterministic code execution. Captions and result notes
state when a table uses the 40-task analysis set, a 25\% subset, a single seed,
or per-arm parse-normalized denominators. Unless otherwise stated, accuracies
are percentages and route differences are percentage points.

\subsection{Complexity-Stratified Route Accuracy}
\label{app:complexity_stratified_route_accuracy}

We stratify the main route-accuracy evaluation by the asymptotic complexity
class of each task. Across classes, \RouteThree{} remains the main source of
improvement over \RouteTwo{}, while \RouteTwo{} stays close to \RouteOne{}.

\begin{appendixtableblock}
\footnotesize
\setlength{\tabcolsep}{4pt}
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lrrrrrrrr@{}}
\toprule
Complexity & Tasks & Inst. & R1 & R2 & R3 & R2--R1 & $p_{2,1}$ & R3--R2 \\
\midrule
$O(1)$ & 3 & 180 & 46.2 & 47.3 & 86.2 & 1.0 & 0.0206 & 39.0 \\
$O(\log n)$ & 1 & 35 & 22.4 & 28.6 & 32.9 & 6.2 & $2.7{\times}10^{-4}$ & 4.3 \\
$O(n)$ & 4 & 135 & 4.8 & 5.0 & 7.5 & 0.2 & 0.635 & 2.5 \\
$O(n \log n)$ & 5 & 68 & 0.0 & 0.0 & 0.0 & 0.0 & 1 & 0.0 \\
$O(n^2)$ & 16 & 297 & 10.3 & 10.5 & 38.9 & 0.1 & 0.644 & 28.5 \\
$O(n^2 \log n)$ & 1 & 1 & 0.0 & 0.0 & 0.0 & 0.0 & 1 & 0.0 \\
$O(n^3)$ & 4 & 37 & 0.0 & 0.0 & 0.0 & 0.0 & 1 & 0.0 \\
NP-hard & 6 & 360 & 17.6 & 16.8 & 69.7 & -0.8 & 0.0278 & 53.0 \\
\bottomrule
\end{tabular*}
\captionof{table}{Route accuracy grouped by asymptotic complexity on the 40-task analysis set in the main route evaluation. Tasks and Inst. report task and instance counts; $p_{2,1}$ is the exact two-sided McNemar test for \RouteTwo{} vs. \RouteOne{} on paired outcomes.}
\label{tab:appendix_complexity_routes}
\end{appendixtableblock}

\appendixresult{\Cref{tab:appendix_complexity_routes} shows the aggregate route pattern within the higher-support complexity strata: \RouteThree{}--\RouteTwo{} is the largest observed route change, while \RouteTwo{} stays close to \RouteOne{}.}

\noindent\textbf{Task mapping.}~The 40-task analysis set excludes
\texttt{segments\_intersect}, \texttt{knap}, \texttt{gcp}, and \texttt{spp}
from the broader 44-task route-analysis mapping. The largest retained strata
are $O(n^2)$ dynamic-programming, sorting, string, graph, and shortest-path
tasks; and NP-hard ILP, edge-disjoint paths, shortest-path, and TSP tasks.
The smaller strata cover constant-time arithmetic, logarithmic binary search,
linear selection/string matching,
$O(n \log n)$ scheduling/sorting/hulls, $O(n^2 \log n)$ Kruskal MST, and cubic
dynamic-programming/shortest-path routines.

\subsection{Model-Stratified Route Accuracy}
\label{app:model_stratified_route_accuracy}

We stratify the main route comparison by model to check whether the aggregate
result is driven by one model family or scale regime. The \RouteThree{} advantage
persists across all six full-coverage models, while the \RouteTwo{}--\RouteOne{}
difference is small and changes sign across models.

\begin{appendixtableblock}
\footnotesize
\setlength{\tabcolsep}{2pt}
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}>{\raggedright\arraybackslash}p{0.36\linewidth}lrrrrrrrr@{}}
\toprule
Model & Type & Inst. & R1 & R2 & R3 & R2--R1 & $p_{2,1}$ & R3--R2 & $p_{3,2}$ \\
\midrule
\path{anthropic/claude-haiku-4.5} & closed & 1113 & 23.3 & 23.6 & 47.1 & 0.3 & 0.612 & 23.5 & $2.4{\times}10^{-148}$ \\
\path{google/gemini-2.0-flash-001} & closed & 1113 & 17.5 & 18.1 & 51.8 & 0.7 & 0.137 & 33.7 & $8.5{\times}10^{-298}$ \\
\path{google/gemini-2.5-flash} & closed & 1113 & 20.7 & 20.5 & 50.2 & -0.1 & 0.823 & 29.7 & $3.1{\times}10^{-238}$ \\
\path{openai/gpt-4o-mini} & closed & 1113 & 14.1 & 12.9 & 49.1 & -1.2 & 0.00663 & 36.1 & $<10^{-300}$ \\
\path{mistralai/codestral-2508} & open & 1113 & 15.0 & 15.8 & 52.7 & 0.8 & 0.0279 & 36.9 & $<10^{-300}$ \\
\path{mistralai/mixtral-8x22b-instruct} & open & 1113 & 12.7 & 13.2 & 42.1 & 0.5 & 0.127 & 28.9 & $7.2{\times}10^{-253}$ \\
\bottomrule
\end{tabular*}
\captionof{table}{Route accuracy grouped by model on the 40-task analysis set. Each row contains 1,113 unique problems and 3,339 paired route evaluations; $p_{2,1}$ and $p_{3,2}$ are exact two-sided McNemar tests for \RouteTwo{} vs. \RouteOne{} and \RouteThree{} vs. \RouteTwo{}, respectively.}
\label{tab:appendix_model_routes}
\end{appendixtableblock}

\appendixresult{\Cref{tab:appendix_model_routes} shows that \RouteThree{} stays above \RouteTwo{} for every full-coverage model row, so the aggregate execution gain is not driven by a single weak or strong model.}

\section{Functional Similarity Checks}
\label{app:functional_similarity_checks}

\subsection{Prompt Translation Shot Ablation}
\label{app:prompt_translation_shot_ablation}

The functional similarity test depends on translating code traces back into
natural language. To test sensitivity to prompting, we vary the number of
in-context examples used by the translator. More examples improve
translated-trace utility and narrow the gap to native natural-language traces.

\begin{appendixtableblock}
\small
\setlength{\tabcolsep}{4pt}
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lrrrrrr@{}}
\toprule
Shots & $x$ & $x$ + native NL & $x$ + translated NL & $\Delta$ native & $\Delta$ translated & Gap \\
\midrule
0 & 32.70\% & 55.70\% & 42.41\% & +23.00pp & +9.70pp & -13.29pp \\
1 & 32.70\% & 55.70\% & 45.78\% & +23.00pp & +13.08pp & -9.92pp \\
2 & 32.70\% & 55.70\% & 49.58\% & +23.00pp & +16.88pp & -6.12pp \\
3 & 32.70\% & 55.70\% & 48.52\% & +23.00pp & +15.82pp & -7.17pp \\
4 & 32.70\% & 55.70\% & 49.37\% & +23.00pp & +16.67pp & -6.33pp \\
5 & 32.70\% & 55.70\% & 50.00\% & +23.00pp & +17.30pp & -5.70pp \\
10 (reference) & 39.00\% & 56.50\% & 52.00\% & +17.50pp & +13.00pp & -4.50pp \\
\bottomrule
\end{tabular*}
\captionof{table}{Translation-additivity shot ablation for Claude Haiku 4.5. Gap is translated NL minus native NL. Rows 0--5 use a matched 25\% subset sweep; the 10-shot row is an original main-evaluation reference and is not from the same subset sweep.}
\label{tab:appendix_shot_ablation}
\end{appendixtableblock}

\appendixresult{\Cref{tab:appendix_shot_ablation} shows that translated traces improve over the question-only baseline at every reported shot setting, but remain below native natural-language traces in every row.}

\section{Additional Model Evaluations}
\label{app:additional_model_evaluations}

\subsection{Recursive Language Model Evaluation}
\label{app:recursive_language_model_evaluation}

To broaden coverage beyond standard LLM forward passes, we evaluate recursive
language model execution on a 25\% subset at seed~0. The code and
natural-language arms remain close under this executor family.

\begin{appendixtableblock}
\small
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lrrr@{}}
\toprule
Model & RLM Code Acc & RLM NL Acc & Better Arm \\
\midrule
Claude Haiku 4.5 & 29.4\% & 30.9\% & NL \\
Gemini 2.5 Pro & 47.5\% & 45.2\% & Code \\
\bottomrule
\end{tabular*}
\captionof{table}{Recursive language model code-vs.-NL accuracy on the 25\% subset, seed~0. Better Arm marks the higher row accuracy.}
\label{tab:appendix_rlm}
\end{appendixtableblock}

\appendixresult{\Cref{tab:appendix_rlm} has no uniform winner: the NL arm is higher for Claude Haiku 4.5, while the code arm is higher for Gemini 2.5 Pro. The corresponding counts are 202/686 code-correct vs. 212/686 NL-correct for Claude Haiku 4.5, and 326/686 code-correct vs. 310/686 NL-correct for Gemini 2.5 Pro. Because this is a 25\% subset, we use it as model-coverage evidence rather than as the main route estimate.}

\subsection{Coding-Specialized Model Evaluation}
\label{app:coding_specialized_model_evaluation}

We evaluate coding-specialized models to test whether models trained on code
change the \RouteTwo{}--\RouteOne{} pattern. Even for these models, replacing
natural-language reasoning with simulation over code traces changes little
relative to replacing simulation with actual code execution.

\begin{appendixtableblock}
\footnotesize
\setlength{\tabcolsep}{4pt}
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}>{\raggedright\arraybackslash}p{0.39\linewidth}ccc@{}}
\toprule
Model & NL & Sim & Code Exec \\
\midrule
\path{x-ai/grok-code-fast-1} (25\% data) & 47.71\% (167/350) & 47.71\% (167/350) & 55.99\% (159/284) \\
\path{qwen/qwen3-coder} (25\% data) & 30.23\% (104/344) & 26.86\% (94/350) & 56.19\% (168/299) \\
\path{codestral-2508} (original) & 19.89\% (943/4740) & 23.14\% (1097/4740) & 59.65\% (2266/3799) \\
\bottomrule
\end{tabular*}
\captionof{table}{Route accuracy for coding-specialized models. NL, Sim, and Code Exec correspond to \RouteOne{}, \RouteTwo{}, and \RouteThree{}; each cell reports accuracy with correct/denominator counts, and denominators are parse-normalized per arm.}
\label{tab:appendix_coding_models}
\end{appendixtableblock}

\appendixresult{\Cref{tab:appendix_coding_models} shows Code Exec as the highest-accuracy arm in each coding-specialized row. Grok and Qwen use 25\% subset runs; Codestral uses the original evaluation run. Rows use per-arm parse-normalized denominators and are not pooled into the main six-model estimate.}

\subsection{Frontier Model Controlled-Subset Evaluation}
\label{app:frontier_model_sanity_checks}

We also evaluate two frontier models on a 350-instance subset at seed~1 with
tool calling disabled and no patching or repair step. These results are not used
as the main estimate because they are smaller than the full six-model
evaluation, but they provide a controlled check that the route pattern is not
explained only by explicit tool calls.

\begin{appendixtableblock}
\small
\setlength{\tabcolsep}{5pt}
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lrrr@{}}
\toprule
Model & Route 1 & Route 2 & Route 3 \\
\midrule
GPT-5.4 & 42.57\% (149/350) & 41.43\% (145/350) & 54.01\% (175/324) \\
Claude Opus 4.6 & 52.73\% (174/330) & 73.42\% (116/158) & 77.54\% (107/138) \\
\bottomrule
\end{tabular*}
\captionof{table}{Frontier tool-disabled controlled-subset route accuracy on the seed~1 350-instance subset with no patching or repair. Entries report accuracy with correct/denominator counts.}
\label{tab:appendix_frontier_nopatch}
\end{appendixtableblock}

\appendixresult{\Cref{tab:appendix_frontier_nopatch} keeps \RouteThree{} ahead in both tool-disabled subset rows. Reasoning-mode runs are excluded because hidden reasoning and tool scaffolding would confound the route comparison.}

\section{Failure Analysis}
\label{app:failure_analysis}

\subsection{Code Execution Failure Modes}
\label{app:code_execution_failure_modes}

Among \RouteThree{} code-execution failures, most failures are semantic wrong
answers rather than syntax or timeout failures. This supports the interpretation
that these errors often originate in the generated program specification rather
than in the Python runtime itself.

\begin{appendixtableblock}
\small
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lrrrr@{}}
\toprule
Model & Wrong answer & Syntax error & Runtime error & Time limit \\
\midrule
Haiku & 56.26\% & 10.07\% & 1.88\% & 31.80\% \\
Codestral & 81.80\% & 5.87\% & 9.00\% & 3.33\% \\
Gemini 2.0 & 77.22\% & 14.65\% & 1.86\% & 6.27\% \\
Gemini 2.5 & 77.63\% & 17.10\% & 1.82\% & 3.45\% \\
GPT-4o mini & 72.93\% & 4.56\% & 18.97\% & 3.54\% \\
Mixtral & 60.67\% & 4.95\% & 32.14\% & 2.24\% \\
Total & 70.43\% & 9.34\% & 11.55\% & 8.67\% \\
\bottomrule
\end{tabular*}
\captionof{table}{Failure-mode distribution conditional on \RouteThree{} code-execution failures. Rows are within-model percentages and sum to 100\% up to rounding.}
\label{tab:appendix_code_failure_modes}
\end{appendixtableblock}

\appendixresult{\Cref{tab:appendix_code_failure_modes} is conditional on \RouteThree{} code-execution failures and excludes parse-only rows where execution completed but the returned value failed parsing or type checks. This keeps the diagnostic focused on what happened after code execution was actually attempted.}
